---
title: "Intolerance In-Class Activity Solutions"
output: pdf_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggrepel)
```

## Step 1: Load data

```{r}
gene_df <- read_tsv("gnomAD_gene_counts_subset_RVIS_upd.tsv")
head(gene_df)
```

```{r}
## generate a scatterplot of the data
gene_df_filt <- gene_df %>% filter(num_common_functional<140)
ggplot(data=gene_df_filt, mapping=aes(x=total_coding_variants, y=num_common_functional)) +
  geom_point() +
  xlab("Total Num. Protein-Coding Variants") +
  ylab("Num. Common (MAF>0.1%) Functional Variants") +
  theme_bw() +
  ggtitle("Total Protein-Coding Variants vs Common Functional Variants") 
```

## Step 2: fit the linear model and calculate studentized residuals

```{r}
rvis_regress <- lm(num_common_functional ~ total_coding_variants, data=gene_df)
summary(rvis_regress)
```

```{r}
student_resid <- rstudent(rvis_regress)
head(student_resid)
```

```{r}
gene_df$RVIS = student_resid
```

```{r}
## to create a plot like the one in the paper
gene_df <- gene_df %>%
  mutate(RVIS_group = case_when(RVIS <= quantile(RVIS, 0.02) ~ "Bottom 2% (Most Intolerant)",
                                RVIS >= quantile(RVIS, 0.98) ~ "Top 2% (Least Intolerant)",
                                TRUE ~ ""))

gene_df_2 <- gene_df %>% filter(num_common_functional<140)

label_df <- bind_rows(
  slice_max(gene_df_2, RVIS, n = 5),
  slice_min(gene_df_2, RVIS, n = 5)
)

ggplot(gene_df, aes(y=num_common_functional, x=total_coding_variants)) +
  geom_point(color = "grey85", size = 2) +
  geom_point(
    data = filter(gene_df, RVIS_group != ""),
    aes(color = RVIS_group),
    size = 2.5
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_color_manual(
    values = c(
      "Bottom 2% (Most Intolerant)" = "red",
      "Top 2% (Least Intolerant)" = "blue"
    )
  ) +
  theme_bw() +
  geom_text_repel(data = label_df, aes(label = gene), size = 3) +
  xlab("Total Num. Protein-Coding Variants") +
  ylab("Num. Common (MAF>0.1%) Functional Variants") +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  ylim(c(0,140)) +
  xlim(c(0,20000))
```

## Step 3: examine a few genes and interpret

```{r}
extremes_df <- bind_rows(
  slice_max(gene_df, RVIS, n = 10),
  slice_min(gene_df, RVIS, n = 10)
)
print(extremes_df)
```

```{r}
gene_df <- gene_df %>% mutate(RVIS_percentile = percent_rank(RVIS)*100)
```

```{r}
## disease gene associated with developmental epileptic encephalopathy
gene_df %>% filter(gene=="SCN1A")
```

```{r}
## disease gene associated with Kabuki syndrome
gene_df %>% filter(gene=="KMT2D")
```

```{r}
gene_df %>% filter(gene=="SETD1A")
```

```{r}
gene_df %>% filter(gene=="ARID1B")
```

```{r}
gene_df %>% filter(gene=="POGZ")
```

```{r}
gene_df %>% filter(gene=="OR10G4")
```

```{r}
gene_df %>% filter(gene=="TAS1R2")
```

```{r}
gene_df %>% filter(gene=="OR1A1")
```

| Gene | RVIS | RVIS Interpreted | LOEUF | LOEUF Interpreted | Function/ Disease Association |
|----------|---------|------------|--------|-------------|---------------------|
| SCN1A | -1.2511 | RVIS \< 0 and in lowest 10%, strong evidence of intolerance | 0.11 | LOEUF \< 0.6 Strong evidence of intolerance | Dravet syndrome, epilepsy, developmental & epileptic encephalopathy |
| SETD1A | -3.6086 | RVIS \< 0 and in lowest 1%, very strong evidence of intolerance | 0.25 | LOEUF\<0.6, strong evidence of intolerance | Neurodevelopmental disorder with speech impairment, epilepsy |
| KMT2D | -5.3856 | RVIS\<0 and in lowest 1%, very strong evidence of intolerance | 0.13 | LOEUF\<0.6, strong evidence of intolerance | Kabuki syndrome |
| ARID1B | -3.9339 | RVIS\<0 and in lowest 1%, very strong evidence of intolerance | 0.19 | LOEUF\<0.6, strong evidence of intolerance | Coffin-Siris syndrome |
| POGZ | -1.5874 | RVIS\<0 and in lowest 5%, strong evidence of intolerance | 0.19 | LOEUF\<0.6, strong evidence of intolerance | White-Sutton syndrome, ASD |
| OR10G4 | 2.6786 | RVIS\>0 and in top 90%, evidence of not intolerant | N/A | N/A | Olfactory receptor gene |
| TAS1R2 | 1.0424 | RVIS\>0 and in top 90%, evidence of not intolerant | 1.21 | LOEUF\>1, evidence of not intolerant | Taste receptor gene, contributes to sensory perception of sweet taste |
| OR1A1 | 1.3081 | RVIS\>0 and in top 90%, evidence of not intolerant | N/A | N/A | Olfactory receptor gene |
|  |  |  |  |  |  |
